A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-24 paper/fascd.tex
completed proper write-up of 1D p-lap results, I think

FIXME smoothers including Newton active set \cite{BensonMunson2006} or Newton semi-smooth \cite{BensonMunson2006}, with either direct linear solve or a fixed number of Krylov steps; in latter case either fixed number of CG+ICC or fixed number of GMRES+ILU

@@

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python using the Firedrake library \cite{Rathgeberetal2016}.  In this Section we show numerical results corresponding to several VI examples from Section \ref{sec:vi}, with a focus on the scaling of FASCD iterations as the mesh resolution increases.

\begin{example} \label{ex:results:classical}
Consider the 2D classical, unilateral obstacle problem, namely Example \ref{ex:plaplacian} with $p=2$ and admissible set $\mathcal{K} = \{v \ge \psi\} \subset W^{1,2}(\Omega)$, with $\Omega$ square.  We consider two problems with very different coincidence (active) sets $\{x\in\Omega \,:\, u(x)=\psi(x)\}$.   The first is a standard ``ball'' problem, with an upper hemisphere obstacle $\psi$ \cite[Chapter 12]{Bueler2021}, and the second is the ``spiral'' problem in which $\psi$ is constructed to produce a more interesting coincidence set \cite[Example 7.1.1]{GraeserKornhuber2009}.  For this experiment we generated uniform triangular meshs, from a coarsest mesh of 32 triangles and $N=4^2$ nodes, to a finest mesh of $N=2048^2 \approx 4.2 \times 10^{6}$ nodes after nine standard refinements, and note that the V- and F-cycles descended to the $4\times 4$ mesh.  A symmetric RS smoother (Section \ref{sec:implementation}) with three (fixed) ICC-preconditioned CG iterations was applied, the coarse solver was RS with LU (direct) solution of the Newton step equations, and tolerances \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$ were used in stopping criterion \eqref{eq:stoppingcriterion}.

Table \ref{tab:results:classical} shows the resulting iterations.  The RS Newton method, i.e.~the smoother, applied by itself as the solver, shows rapidly increasing iterations, approximately proportionally to resolution, as expected.  By contrast, FASCD V-cycle iterations increase slowly, roughly logarithmically with the resolution.  For F-cycles we counted the number of V-cycles after the ramp (Section \ref{sec:implementation}), and this shows a constant number of V-cycle iterations.  The spiral problem appears to be only slightly more challenging for these solvers; apparently the complexity of the coincidence set or free boundary does not strongly influence performance.

Note that each V-cycle does the work of about eight finest-level preconditioned Krylov (CG+ICC) iterations.  For the F-cycle results, where three V-cycle iterations suffice in either problem, and at any finest-level resolution, we see textbook multigrid efficiency\footnote{This phrase is imperfectly defined, but one possible definition is given by \cite{BrownSmithAhmadia2013}: ``textbook multigrid efficiency [is] (an order of magnitude reduction in residual per iteration and solution of the fine-level system at a small multiple of the cost of a residual evaluation).''} despite their low solution regularity.
\end{example}

%REGENERATE using pefarrell-fascd/examples/
%$ timer python3 drape.py -prob ball|spiral -levs 9 -o ball|spiral.pvd -fascd_rtol 1.0e-6 -fascd_atol 1.0e-6 -fascd_cycle_type full
%$ paraview ball|spiral.pvd
%use preset "xray" and invert colors
\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.35\textwidth]{fixfigs/ball-set.png} \qquad \includegraphics[width=0.35\textwidth]{fixfigs/spiral-set.png}
\end{center}
\caption{Coincidence sets for the ball and spiral problems in Example \ref{ex:results:classical}.}
\label{fig:ballspiral}
\end{figure}

\newcommand{\XX}{{\sffamily X}}
% see pefarrell/examples/drape.py and pefarrell/edplay/resultsdrape.sh|txt
% note mcoarse=4, rtol=1e-6, atol=1e-12
% smoother is vinewtonrsls + 3 x (cg + icc), coarse solver is vinewtonrsls + lu
\begin{table}[ht]
\begin{tabular}{lr rrr rrr}
\multirow{2}{*}{\emph{levels}} & \multirow{2}{*}{$N$} & \multicolumn{3}{c}{\,\emph{ball}} & \multicolumn{3}{c}{\,\emph{spiral}} \\ \cmidrule(lr){3-5} \cmidrule(lr){6-8}
             && \, V-cycle & F-cycle & RS only & \,V-cycle & F-cycle & RS only \\ \cmidrule{1-8}
 1 &    $4^2$ &  1 &  1 &  1  &  1 &  1 &  1 \\
 2 &    $8^2$ &  2 &  1 &  2  &  2 &  1 &  2 \\
 3 &   $16^2$ &  2 &  2 &  3  &  3 &  2 &  4 \\
 4 &   $32^2$ &  3 &  2 &  6  &  3 &  2 &  7 \\
 5 &   $64^2$ &  3 &  2 &  8  &  4 &  3 &  9 \\
 6 &  $128^2$ &  4 &  2 & 17  &  4 &  3 & 16 \\
 7 &  $256^2$ &  4 &  2 & 33  &  4 &  3 & 27 \\
 8 &  $512^2$ &  5 &  2 & 65  &  5 &  3 & 48 \\
 9 & $1024^2$ &  5 &  2 & \XX &  5 &  3 & \XX \\
10 & $2048^2$ &  5 &  2 & \XX &  5 &  3 & \XX
\end{tabular}

\bigskip
\caption{Iterations for FASCD V-cycle and F-cycle algorithms in Example \ref{ex:results:classical}.  Runs marked \XX\xspace were not attempted.}
\label{tab:results:classical}
\end{table}

\begin{example}  \label{ex:results:plap}
FIXME results for 1D $p$-Laplacian obstacle problem, Example \ref{ex:plaplacian} with $p=1.5,4$; this is a good problem to visualize the problems solved on each coarser level Figure \ref{fig:imagesvcycle}
\end{example}

% figures in the following visualization were generated using branch elbueler/simplefigs
% of git@bitbucket.org:pefarrell/fascd.git, and the following commands:
%   cd examples/
%   python3 plap1d.py -figures -fascd_max_it 2 -fascd_cycle_type multiplicative -mcoarse 2
\begin{figure}[ht]
\begin{center}
\input{tikz/imagesvcycle.tex}
\end{center}
\caption{Visualization of one FASCD V-cycle iteration for Example \ref{ex:results:plap}, with $p=1.5$, $J=3$, and a coarsest mesh of two elements.  Solutions (i.e.~$w^j$, $y^j$, $z^j$) are lines while obstacles ($\underline{\gamma}^j$, $\underline{\phi}^j$, $\underline{\chi}^j$) are strong dots.  The zero function (dotted) is the admissible initial iterate for $y^j$ and $z^j$.  Vertical scale varies.}
\label{fig:imagesvcycle}
\end{figure}

\begin{example}  \label{ex:results:advdiff}
FIXME results for advection-diffusion obstacle problem, Example \ref{ex:advectiondiffusion}
\end{example}

>>>>

FIXME smoothers use Newton reduced space (RS) \cite{BensonMunson2006}with either direct linear solve or a fixed number of Krylov steps; in latter case either fixed number of CG+ICC for symmetric linear operator (Laplacian), otherwise fixed number of GMRES+ILU

% FIXME try Newton semi-smooth \cite{BensonMunson2006}?

@@

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python using the Firedrake library \cite{Rathgeberetal2016}.  In this Section we show numerical results corresponding to several VI examples from Section \ref{sec:vi}, with a focus on the scaling of FASCD iterations as the mesh resolution increases.  Our essential observation is that, with strong $O(m_J)$ smoothers on each level, each V-cycle dramatically reduces the residual norm, and generally a couple of F-cycles solves the problem down to discretization error.

\begin{example} \label{ex:results:classical}
Consider the 2D classical, unilateral obstacle problem, namely Example \ref{ex:plaplacian} with $p=2$ and admissible set $\mathcal{K} = \{v \ge \psi\} \subset W^{1,2}(\Omega)$, with $\Omega$ square.  We consider two problems with very different coincidence (active) sets $\{x\in\Omega \,:\, u(x)=\psi(x)\}$.   The first is a standard ``ball'' problem, with an upper hemisphere obstacle $\psi$ \cite[Chapter 12]{Bueler2021}, and the second is the ``spiral'' problem from Example 7.1.1 in \cite{GraeserKornhuber2009}, wherein $\psi$ is constructed to produce a more interesting coincidence set; see Figure \ref{fig:results:classical}.  For this experiment we used $P_1$ elements and uniform triangular meshs, from a coarsest mesh of $m_0=5^2$ nodes (32 triangles), to a finest mesh of $m_9=2049^2 \approx 4.2 \times 10^{6}$ nodes after standard refinements.  As our $O(m_j)$ smoother we applied a single iteration of RS Newton (Section \ref{sec:implementation}), with three (fixed) ICC-preconditioned CG iterations to approximately solve the step equations.  The coarse solver was RS$+$LU, iterated to convergence.  Tolerances \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$ were used in stopping criterion \eqref{eq:stoppingcriterion}.  Noting that $\psi\big|_{\partial \Omega} \le 0$ in both problems, the initial iterate was set to $w=\max\{0,\psi\}$.

Table \ref{tab:results:classical} shows the resulting iteration counts.  When the RS Newton smoother is applied by itself as a solver it shows rapidly-increasing iterations as expected, approximately proportionally to resolution.  By contrast, FASCD V-cycle iterations increase slowly, roughly logarithmically with the resolution.  Noting that for F-cycles we are counting the number of V-cycles after the ramp (Section \ref{sec:implementation}), we see that a constant number of F-cycles solves the problem.  The spiral problem appears to be only slightly more challenging for these solvers; apparently the complexity of the coincidence set or free boundary does not strongly influence performance.

Note that each V-cycle does the work of about eight finest-level preconditioned Krylov (CG+ICC) iterations.  For the F-cycle results, where three V-cycle iterations suffice in either problem, and at any finest-level resolution, this is nearly textbook multigrid efficiency\footnote{Textbook multigrid efficiency is imperfectly defined, but consider \cite{BrownSmithAhmadia2013}: ``an order of magnitude reduction in residual per iteration and solution of the fine-level system at a small multiple of the cost of a residual evaluation.''  However, work of at most ten residual evaluations is often mentioned.} despite the low solution regularity.
\end{example}

%REGENERATE using git@bitbucket.org:pefarrell/fascd.git, in examples/
%$ timer python3 drape.py -prob ball|spiral -levs 9 -o ball|spiral.pvd -fascd_rtol 1.0e-6 -fascd_atol 1.0e-6 -fascd_cycle_type full
%$ paraview ball|spiral.pvd
%use preset "xray" and invert colors
\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.35\textwidth]{fixfigs/ball-set.png} \qquad \includegraphics[width=0.35\textwidth]{fixfigs/spiral-set.png}
\end{center}
\caption{Coincidence sets for the ball and spiral problems in Example \ref{ex:results:classical}.}
\label{fig:results:classical}
\end{figure}

% see git@bitbucket.org:pefarrell/fascd.git: examples/drape.py and edplay/resultsdrape.sh|txt
% note mcoarse=4, rtol=1e-6, atol=1e-12
% smoother is vinewtonrsls + 3 x (cg + icc), coarse solver is vinewtonrsls + lu
\begin{table}[ht]
\begin{tabular}{lr@{\hskip 7mm}cccccc}
\multirow{2}{*}{\emph{levels}} & \multirow{2}{*}{$m_J$} & \multicolumn{3}{c}{\,\emph{ball}} & \multicolumn{3}{c}{\,\emph{spiral}} \\ \cmidrule(lr){3-5} \cmidrule(lr){6-8}
   &          & V-cycle & F-cycle & RS only & V-cycle & F-cycle & RS only \\ \cmidrule{1-8}
 1 &    $5^2$ &  1 &  1 &  1  &  1 &  1 &  1 \\
 2 &    $9^2$ &  2 &  1 &  2  &  2 &  1 &  2 \\
 3 &   $17^2$ &  2 &  2 &  3  &  3 &  2 &  4 \\
 4 &   $33^2$ &  3 &  2 &  6  &  3 &  2 &  7 \\
 5 &   $65^2$ &  3 &  2 &  8  &  4 &  3 &  9 \\
 6 &  $129^2$ &  4 &  2 & 17  &  4 &  3 & 16 \\
 7 &  $257^2$ &  4 &  2 & 33  &  4 &  3 & 27 \\
 8 &  $513^2$ &  5 &  2 & 65  &  5 &  3 & 48 \\
 9 & $1025^2$ &  5 &  2 & \XX &  5 &  3 & \XX \\
10 & $2049^2$ &  5 &  2 & \XX &  5 &  3 & \XX
\end{tabular}
\bigskip
\caption{Iterations for FASCD V-cycle and F-cycle algorithms in Example \ref{ex:results:classical}.  Smoother-only runs marked \XX\xspace were not attempted.}
\label{tab:results:classical}
\end{table}

\begin{example}  \label{ex:results:plap}
Recall Example \ref{ex:plaplacian}, the nonlinear $p$-Laplacian operator.  We will solve a unilateral obstacle problem, in one dimension for visualizability, considering both fast ($p=1.5$) and degenerate diffusivity ($p=4.0$) cases.\footnote{``Fast'' and ``degenerate'' refer to $D=|\grad u|^{p-2}\to \infty$ and $D\to 0$ as $|\grad u|\to 0$, respectively.  Ellipticity is lost in the latter case.}  This example reveals a strong contrast between $V(1,0)$ and $V(0,1)$ results, namely \pr{up} $=0$ and \pr{down} $=0$ in Algorithm \ref{alg:fascd}, respectively, and it emphasizes the need for robust smoothers.

Suppose $\Omega=(-3,3)$, $\psi(x) = -0.2|x|$, $\mathcal{K} = \{v \ge \psi\} \subset W^{1,p}(\Omega)$, and $\ell(v) = \int_\Omega g v$ where $g(x)=1$ for $|x|<1$ and $g(x)=-1$ for $|x|>1$.  The exact solution of VI \eqref{eq:vi} is easily computed for any $p$; see Figure \ref{fig:results:plap}.  For this experiment we used $P_1$ elements, a coarsest mesh of $m_0=7$ nodes, and nine standard refinements to a $m_9=3073$ finest mesh.  The default smoother uses three (fixed) iterations of RS Newton, with LU solution of the step equations.\footnote{This is an optimal $O(m_j)$ smoother in 1D.}  The coarse solver is the same but iterated to convergence.  The initial iterate was $w=\psi$, and again \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$.

%REGENERATE using elbueler/plapfigs branch of git@bitbucket.org:pefarrell/fascd.git, in examples/
% $ python3 plap1d.py -p 1.5 -figures -levs 6 -fascd_cycle_type full && mv solution.pdf plap1d1p5.pdf
% $ python3 plap1d.py -p 4.0 -figures -levs 6 -fascd_cycle_type full && mv solution.pdf plap1d4p0.pdf
\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.45\textwidth]{fixfigs/plap1d1p5.pdf} \quad
\includegraphics[width=0.45\textwidth]{fixfigs/plap1d4p0.pdf}
\end{center}
\caption{Solutions $u\ge \psi$ of the $p$-Laplacian VI in Example \ref{ex:results:plap}, for $p=1.5$ (left) and $p=4.0$ (right).}
\label{fig:results:plap}
\end{figure}

Figure \ref{fig:imagesvcycle} illustrates the major steps of an example V-cycle, showing each of the nonlinear FAS coarse correction VI subproblem solutions.  That is, each $y^j \in \mathcal{D}^j$ and $z^j \in \mathcal{U}^j$ is drawn, along with each level defect constraint $\underline{\phi}^j,\underline{\chi}^j$.

% figures in the following visualization were generated using branch elbueler/simplefigs
% of git@bitbucket.org:pefarrell/fascd.git, and the following commands:
%   cd examples/
%   python3 plap1d.py -figures -fascd_max_it 2 -fascd_cycle_type multiplicative -mcoarse 2
\begin{figure}[ht]
\begin{center}
\input{tikz/imagesvcycle.tex}
\end{center}
\caption{Visualization of one FASCD V-cycle iteration for Example \ref{ex:results:plap}, with $p=1.5$, $J=3$, and a coarsest mesh of $m_0=3$ nodes.  Solutions (i.e.~$w^j$, $y^j$, $z^j$) are lines while obstacles ($\underline{\gamma}^j$, $\underline{\phi}^j$, $\underline{\chi}^j$) are strong dots.  The zero function (dotted) is the admissible initial iterate for $y^j$ and $z^j$.  Vertical scale varies.}
\label{fig:imagesvcycle}
\end{figure}

Iteration results for the fast diffusion $p=1.5$ case are shown in Table \ref{tab:results:fastplap1d}.  We see that $V(1,0)$ cycles, which correspond to an FAS implemention of Algorithm 4.7 in \cite{GraeserKornhuber2009}, but with a stronger smoother, often do not converge.  By contrast, $V(0,1)$ cycles, using only up-smoothing in the larger sets $\mathcal{U}^j$ (Section \ref{sec:cdmultilevel}), are just as good as the default $V(1,1)$ cycles of Algorithm \ref{alg:fascd}.  These default V-cycles give convergence in the $\|\cdot\|_\infty$ norm (last column) at rate $O(h^{2.005})$.

Turning to the F-cycle results in Table \ref{tab:results:fastplap1d}, they are very efficient except when they fail entirely.  This apparently arises from singular diffusivity.  If we regularize the diffusivity formula, from $D=|\grad u|^{-0.5}$ to $D_\eps = \left(\eps + |\grad u|^2\right)^{-0.25}$ with $\eps=10^{-8}$, then a single F-cycle suffices at every resolution (column ``F$_\eps$'').

Iteration results for the (unregularized) degenerate diffusion $p=4.0$ case are in Table \ref{tab:results:degenerateplap1d}.  V-cycles using the default settings, specifically the fixed three RS Newton iterations, do not converge on the finer meshes; the semi-smooth NCP residual norms \eqref{eq:stoppingcriterion} are unbounded in the failed runs.  Visualization shows an instability in which oscillations in the restricted residuals explode.  Fortunately, either a slight increase in smoother power, to four RS Newton iterations, or improved finest-level initial iterates, through use of an F-cycle, stops the instability.  In fact, a single F-cycle suffices at every resolution, giving convergence at rate $O(h^{1.420})$.
\end{example}

% for Tables \ref{tab:results:fastplap1d} and \ref{tab:results:degenerateplap1d},
% see git@bitbucket.org:pefarrell/fascd.git: examples/plap1d.py and edplay/resultsplap1d.sh|txt

\begin{table}[ht]
\begin{tabular}{lr@{\hskip 7mm}c@{\hskip 3mm}c@{\hskip 3mm}c@{\hskip 4mm}c@{\hskip 5mm}c@{\hskip 6mm}c}
\emph{levels} & $m_J$ & V(1,0) & V(0,1) & V(1,1) & F & F$_\eps$ & $\|$error$\|_\infty$ \\ \cmidrule{1-8}
 1 &    $7$ &  1 &  1 &  1 &  1 &  1 & $2.3\times 10^{-1}$ \\
 2 &   $13$ &  3 &  3 &  2 &  2 &  1 & $3.3 \times 10^{-2}$ \\
 3 &   $25$ & 12 &  4 &  2 &  1 &  1 & $9.1 \times 10^{-3}$ \\
 4 &   $49$ & NC &  4 &  2 &  1 &  1 & $3.2 \times 10^{-3}$ \\
 5 &   $97$ & NC &  3 &  3 &  1 &  1 & $5.5 \times 10^{-4}$ \\
 6 &  $193$ & NC &  3 &  3 & NC &  1 & $1.7 \times 10^{-4}$ \\
 7 &  $385$ & NC &  3 &  3 &  1 &  1 & $4.7 \times 10^{-5}$ \\
 8 &  $769$ & NC &  4 &  7 &  1 &  1 & $9.5 \times 10^{-6}$ \\
 9 & $1537$ & NC &  5 &  6 & NC &  1 & $3.6 \times 10^{-6}$ \\
10 & $3073$ & NC & 11 & 19 &  1 &  1 & $4.9 \times 10^{-7}$
\end{tabular}
\bigskip
\caption{Iterations for V- and F-cycles on the $p=1.5$ problem in Example \ref{ex:results:plap}.  NC $=$ did not converge after 50 iterations.  See text regarding F$_\eps$ and error.}
\label{tab:results:fastplap1d}
\end{table}

\begin{table}[ht]
\begin{tabular}{lr@{\hskip 7mm}c@{\hskip 4mm}c@{\hskip 4mm}c@{\hskip 6mm}c}
\emph{levels} & $m_J$ & V & V$_4$ & F & $\|$error$\|_\infty$ \\ \cmidrule{1-6}
 1 &    $7$ &  1 &  1 &  1 & $8.7 \times 10^{-2}$ \\
 2 &   $13$ &  1 &  1 &  1 & $3.8 \times 10^{-2}$ \\
 3 &   $25$ &  2 &  1 &  1 & $1.6 \times 10^{-2}$ \\
 4 &   $49$ &  4 &  2 &  1 & $6.3 \times 10^{-3}$ \\
 5 &   $97$ & NC &  2 &  1 & $2.3 \times 10^{-3}$ \\
 6 &  $193$ & NC &  1 &  1 & $7.0 \times 10^{-4}$ \\
 7 &  $385$ & NC &  2 &  1 & $2.0 \times 10^{-4}$ \\
 8 &  $769$ & NC &  2 &  1 & $1.1 \times 10^{-4}$ \\
 9 & $1537$ & NC &  2 &  1 & $4.0 \times 10^{-5}$ \\
10 & $3073$ & NC &  2 &  1 & $1.5 \times 10^{-5}$
\end{tabular}
\bigskip
\caption{Iterations for V- and F-cycles on the $p=4.0$ problem in Example \ref{ex:results:plap}.  See text regarding V$_4$ and error.}
\label{tab:results:degenerateplap1d}
\end{table}

\begin{example}  \label{ex:results:advdiff}
FIXME results for 3D advection-diffusion obstacle problem, Example \ref{ex:advectiondiffusion}
\end{example}
